Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-05-25 paper/paper.Rmd
completed section 2

This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting (ref FIXME: <https://doi.org/10.1016/j.csda.2017.11.003>).
It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
In this stituation it would therefore seem rather foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e. using other grid cells from the same fields, as if we wanted to predict within a field. The results from this performance assessment would be over-optimitic, and perhaps badly so.
To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
If the model was to be applied to adjacent agricultural districts, i.e. outside the learning sample's spatial domain (*extra-domain* prediction; @brenning2005), it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.

@@

In cross-validation for predictive model assessment, we consider the following formal setting.
We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$. Our goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
\[
\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X))),
\]
where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@milligan2000].
The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and ref FIXME perhaps <https://doi.org/10.1016/j.ijforecast.2006.03.001> for regression and forecasting tasks).

@@

- add a paragraph with more information on spCV using spatial partitionings

A variant of cross-validation is leave-one-out (LOO) cross-validation , which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
Since this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005].
When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, LOO at has been proposed at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005].
Similarly, in the temporal domain, it has been proposed to leave out temporal observational units, as in leave-one-year-out CV [@anderson2005; @brenning2005].

It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), are not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
'Pure' LOO-CV is therefore not recommended for machine-learning model assessment [ref??].

- add a paragraph on spatiotemporal partitioning

>>>>

This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
The purpose of training a model on this sample is to make predictions on other, new *fields* within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
In this stituation it would therefore seem rather foolish to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e. using other grid cells from the same fields, as if we wanted to predict within a field. The results from this performance assessment would be over-optimitic, and perhaps badly so.
To mimic the predictive situation for which the model is trained, we'd rather have to resample at the level of fields, not grid cells [@pena2015].
If the model was to be applied to adjacent agricultural districts, i.e. outside the learning sample's spatial domain (*extra-domain* prediction; @brenning2005), it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.

@@

In cross-validation for predictive model assessment, we consider the following formal setting.
We are interested in predicting a numerical or categorical response $y$ of an object or instance using a feature vector $\mathbf{x} = (x^{(1)}, \ldots, x^{(p)})\in\mathbb{R}^p$ and a model $\hat{f}_\mathcal{L}$ that has been trained on a learning sample $\mathcal{L} = \{(y_i, \mathbf{x}_i),\ i = 1, ..., n\}$. Our goal is to estimate the expected value of the performance of $\hat{f}_\mathcal{L}$,
\[
\mathit{perf(\hat{f}_\mathcal{L})} := E(l(Y,\hat{f}_\mathcal{L}(X))),
\]
where $l$ is a real-valued loss function, and the expected value is with respect the probability distribution of $X$, the features of an instance $(Y,X)$ drawn randomly from the underlying population.
This is referred to as the *actual* or *conditional* performance measure as it is conditional on $L$ [@milligan2000].
The loss function can take a variety of forms such as the misclassification error $I(Y\neq\hat{f}_\mathcal{L}(X))$ in classification, or the squared error $(Y-\hat{f}_\mathcal{L}(X))^2$ in regression, among many other possible measures.
The choice of the performance measure is equally critical as the choice of the estimation procedure, but it is beyond the scope of this contribution to discuss performance metrics for regression and classification (see e.g., @hand1997 for classification, and @hyndman2006 for regression and forecasting tasks).

@@

Approaches based on spatial blocking or grouping require either the construction of spatial zones, or the use of pre-existing spatial structures in the data. Let's refer to these spatial units or zones as $\mathcal{Z_i}$, $1\le i\le n_z$. 
These zones are often constructed to serve as the $k=n_z$ spatial partitions, for example by performing $k$-means clustering of the sample coordinates [@russ2010] or generating the desired number of rectangular blocks. Similarly, the data can be partitioned in feature space instead of geographic space, which has been referred to as environmental blocking [@roberts2017].

When $n_z$ is much larger than the desired number of folds, $k$, then a partitioning can be applied to the zones themselves. In this case, the zone indices $1, \ldots, n_z$ are grouped into $k$ equally sized subsets $\mathcal{I}_1, \ldots, \mathcal{I}_k$.
This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
In the following we will conceptually distinguish between CV *at the group (or block) level*, referring to the latter scenario, from spatial *block (or grouped)* CV, where the blocks or groups themselves define the CV partitions.

A variant of cross-validation is leave-one-out (LOO) cross-validation, which has long been established in geostatistics [@cressie1993], sometimes with a focus on the spatial distribution of LOO error [@willmott2006].
Although this is just a special case of non-spatial CV with $k=n$, it is sometimes also referred to as spatial CV [@willmott2006].
Nevertheless, spatial variants have been proposed that apply an exclusion buffer to the test locations to separate them from the training data [@brenning2005].
One plausible approach for this is to use the range of autocorrelation of model residuals to determine the buffer distance as this ensures independence conditional on the predictors [@brenning2005].
When data is grouped, e.g., due to multi-level sampling designs or the study of spatial objects, it has been proposed to apply LOO at the site level [@martin2008; @kasurak2011] or, in animal movement studies, at the animal level [@anderson2005]; we will refer to this as leave-location-out (LLO) CV.

It should be noted that $k$-fold cross-validation with a large value of $k$, and LOO-CV in particular ($k=n$), is not only very time-consuming since the model has to be trained $k$ times; these models will also be nearly identical since only a tiny fraction of the data is withheld, and therefore estimation bias increases.
'Pure' LOO-CV is therefore not recommended for machine-learning model assessment.

In the purely temporal domain, a special case is to leave out temporal observational units (or time slices; leave-time-out or LTO CV), as in leave-one-year-out CV [@anderson2005; @brenning2005]. 
Cross-validation and hold-out validation strategies for time series have been discussed more extensively in the forecasting literature, considering also the effects of serial autocorrelation [@bergmeir2018].

Turning to prediction tasks with spatiotemporal data, various spatial, temporal, or spatiotemporal partitioning strategies are being used, depending on the specific study objectives. 
While the former two ignore the spatial and temporal dimension of the data, respectively, it has also been proposed to leave out spatiotemporal blocks [@meyer2018] or clusters [@cluto].
